---
title: "Hurricane Data"
author: "Alison Pedraza"
date: '2024-03-19  (part of a project from 2020)'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(drat)
addRepo("geanders")
library(tidyverse)
library(hurricaneexposuredata)
library(magrittr)
library(ggplot2)
library(dplyr)
#require(usmap)
require(viridis)
#library(tmap)
library(maps)  #Used to map the States by county and get county map lat and long

```



```{r echo=FALSE}
htmltools::img(src = knitr::image_uri("C:/Users/aliso/Documents/My Projects/EDA and Maps/Hurricane.jpg"), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:20px;', )
```

<br>

### Project Purpose and Outline:
***

**Purpose:** 

  + To show the total amount of rainfall (in mm) from the Floyd, Allison, and Irene hurricanes and their tracks on a U.S. map.
  
  + To show data cleaning, preprocessing, and to display data on a map.

**Outline:**
  
  + Hurricane data for Floyd (1999), Allison (2001), and Irene (2011)
  
  + Data Sources & libraries
    
  + Data Cleaning and Pre-processing:
  
      + Grouping
      + Filtering
      + Separating values within a column and placing into their own columns
      + Joining dataframes
      + Changing values from characters to numeric 
      + Mutating values within a column
    
    <br>
    
  + Maps of U.S. states which were affected by these hurricanes
  
<br>


### Data Sources:
***
  + Data came from the following libraries :
  
  
      + hurricaneexposuredata
      + maps
      
<br>

  + Dataframes from **hurricaneexposuredata** library:
  
  
      + "rain"
      + "hurr_tracks"
      + "county.fips"
      
<br>

  + Dataframe from **maps** library:
  
      
      + "county.fips"
  
<br>
  
  + Some other libraries used:
  
  
      + usmap
      + ggplot2
        
<br>

<br>

### Data Cleaning and Preprocessing
***

```{r  echo=TRUE, warning=FALSE, message=FALSE, message=FALSE}
data("rain")
data_rain <- as.data.frame(rain)
data("hurr_tracks")
data_tracks <- as.data.frame(hurr_tracks)
```

<br>

<br>


#### Filter out the storms: 

  + Floyd (1999), Allison(2001), and Irene(2011) from the column storm_id, clean and preprocess:

```{r echo=TRUE, warning=FALSE, message=FALSE, message=FALSE}
storm_data <- data_rain %>% filter(storm_id == "Floyd-1999")
Floyd_tracks <- data_tracks %>% filter(storm_id == "Floyd-1999")
irene_data <- data_rain%>%filter(storm_id == "Irene-2011")
irene_tracks <- data_tracks%>%filter(storm_id == "Irene-2011")
storm_data2 <- data_rain %>% filter(storm_id == "Allison-2001")
Allison_tracks <- data_tracks %>% filter(storm_id == "Allison-2001")

## Group all the fip entries by fip number:

Floyd_1999 <- storm_data %>% group_by(fips, storm_id)%>%
  summarise(total_precip = sum(precip), precip_max = sum(precip_max))

Allison_2001 <- storm_data2 %>% group_by(fips, storm_id)%>%
  summarise(total_precip = sum(precip))

irene_2011 <- irene_data %>%  group_by(fips, storm_id)%>%
  summarise(total_precip = sum(precip))


##Change fips to numeric
Floyd_1999$fips <- as.numeric(Floyd_1999$fips)
#is.numeric(Floyd_1999$fips)

Allison_2001$fips <- as.numeric(Allison_2001$fips)
#is.numeric(Allison_2001$fips)

irene_2011$fips <- as.numeric(irene_2011$fips)
#is.numeric(irene_2011$fips)

data("county.fips")
county_fips <- as.data.frame(county.fips)
county_fips%<>%separate("polyname", c("region", "subregion"), sep = ",")

# #Found map data for the united states and counties
states <- map_data("state")
county <- map_data("county")

## Filtered out the states I want from map_data or states. Called it newStates


some.us.states <- c("maine", "vermont", "new hampshire", "massachusetts", "connecticut", "rhode island",
                     "new york", "new jersey", "delaware", "pennsylvania", "north carolina", "south carolina", 
                     "louisiana", "georgia", "florida", "mississippi", "texas", "virginia", "west virginia", 
                     "tennessee", "maryland", "alabama", "arkansas", "missouri", "illinois", "kentucky", "indiana",
                     "ohio", "michigan", "wisconsin", "oklahoma", "kansas", "iowa")


#Get only the states and counties I want - using vector.

county.map <- map_data("county", some.us.states)
Floyd_states <- map_data("state", some.us.states)
Allison_states <- map_data("state", some.us.states)
irene_states <- map_data("state", some.us.states)

# Merge and get lat and long into dataframes Floyd_storm data.
Floyd_storm <- left_join(county_fips, Floyd_1999, by= c("fips"))
Floyd_Rain <- left_join(Floyd_storm, county.map, by = c( "region", "subregion"))
Floyd_Rain<- Floyd_Rain %>% mutate('Rain (mm)' = cut(Floyd_Rain$total_precip, 
                                                     breaks = c(0,25,50,75,100,125,150,175,200,223),
                                                  include.lowest = TRUE))
Floyd_Rain<- na.omit(Floyd_Rain)

Allison_2001 <- left_join(Allison_2001, county_fips, by = "fips")
Allison_2001<- left_join(county.map, Allison_2001, by = c("region", "subregion"))
Allison_2001 <- Allison_2001 %>%
  mutate(`Rain > 175 (mm)` = ifelse(Allison_2001$total_precip < 175, "Unexposed", "Exposed"))
Allison_2001 <- na.omit(Allison_2001)
Allison_Rain<- Allison_2001 %>% mutate('Rain (mm)' = cut(Allison_2001$total_precip, 
                                                     breaks = c(0,50,100,150,200, 250, 300, 350, 400, 440),
                                                     include.lowest = TRUE))
Allison_Rain<- na.omit(Allison_Rain)

irene_2011 <- left_join(irene_2011, county_fips, by = "fips")
irene_2011<- left_join(county.map, irene_2011, by = c("region", "subregion"))
irene_2011 <- irene_2011 %>%
  mutate(`Rain > 175 (mm)` = ifelse(irene_2011$total_precip < 175, "Unexposed", "Exposed"))
irene_2011 <- na.omit(irene_2011)
irene_Rain<- irene_2011 %>% mutate('Rain (mm)' = cut(irene_2011$total_precip, 
                                                     breaks = c(0,50,75, 100, 125, 150, 175, 200, 225, 250, 275, 300, 325),
                                                     include.lowest = TRUE))
irene_Rain<- na.omit(Allison_Rain)
```

<br>

<br>

## Maps
***

<br>

### Floyd map: Total Precipitation and Hurricane track


```{r echo=TRUE, warning=FALSE, message=FALSE, message=FALSE}
ggplot() + geom_polygon(data = Floyd_Rain, aes(x = long, y = lat, group = group, fill = `Rain (mm)`),
                        color = "lightgrey", size = .2, alpha = 1.0) + geom_polygon(data = Floyd_states, aes(x = long, y = lat, group = group), 
                       color = "black", fill = "transparent",  size = 0.1, alpha = 0.3)+
  scale_fill_brewer(palette = "Blues") +
  geom_path(aes(x = Floyd_tracks$longitude, y = Floyd_tracks$latitude), color = "red") +
  ggtitle("Hurricane Floyd - 1999: Total Precipitation and Track")
```

<br>

<br>

### Allison map: Total Precipitation and Hurricane track


```{r echo=TRUE, warning=FALSE, message=FALSE, message=FALSE}
ggplot() + geom_polygon(data = Allison_Rain, aes(x = long, y = lat, group = group, fill = `Rain (mm)`),
                        color = "lightgrey", size = .2, alpha = 1.0) + geom_polygon(data = Allison_states, aes(x = long, y = lat, group = group), 
                       color = "black", fill = "transparent",  size = 0.1, alpha = 0.3) +
  scale_fill_brewer(palette = "Set3") +
  geom_path(aes(x = Allison_tracks$longitude, y = Allison_tracks$latitude), color = "red") +
  ggtitle("Hurricane Allison-2001: Total Precipitation (mm) and Track")
```


<br>

<br>

### Irene map: Total Precipitation and Hurricane track


```{r echo=TRUE, warning=FALSE, message=FALSE, message=FALSE}
ggplot() + geom_polygon(data = irene_Rain, aes(x = long, y = lat, group = group, fill = `Rain (mm)`),
                        color = "lightgrey", size = .2, alpha = 1.0) + geom_polygon(data = irene_states, aes(x = long, y = lat, group = group), 
                       color = "black", fill = "transparent",  size = 0.1, alpha = 0.3) +
  scale_fill_brewer(palette = "YlOrRd") +
  geom_path(aes(x = irene_tracks$longitude, y = irene_tracks$latitude), color = "red") +
  ggtitle("Hurricane Irene-2011: Total Precipitation (mm) and Track")
```

<br>
